Rapid and Accurate Global PTM Discovery (G-PTM-D) Using Post-Acquisition Spectral Calibration and Defined Mass Windows

### 2017-04-19 paper.tex
many

Correct identification of protein post-translational modifications (PTMs) is crucial to understanding many aspects of protein function in biological processes.
G-PTM-D\cite{Li_2016} is a recently developed tool for global identification and localization of PTMs.
Spectral file calibration prior to applying G-PTM-D, and algorithmic enhancements in the peptide database search significantly increase the accuracy and efficiency of PTM identification.
We enhance G-PTM-D by using targeted mass-window searches, and demonstrate its effectiveness in identification of numerous types of PTMs, including high-mass modifications such as glycosylations.
The proposed changes lead to a 20\% increase in the number of identified modifications and an order of magnitude decrease in the search time.
Targeted mass-window searches are also shown to be useful as replacements to standard wide-mass searches, resulting in a 10\% increase in the overall number of confidently identified peptides, and a 5\% increase in the number of mass shifts observed.
Using a modification discovery heuristic originally based on the wide-mass search, targeted mass-window searches lead to an increase in the number of novel modifications discovered.
\end{abstract}

@@

\begin{figure}
  \includegraphics[scale=0.5]{diagram.eps}
  \caption{Workflows}
  \label{fgr:diagram}
\end{figure}

@@

\begin{figure}
  \includegraphics[scale=0.85]{fig2-calibrationQuality.eps}
  \caption{Histograms demonstrating the effect of calibration on PSM mass errors}
  \label{fgr:figure1}
\end{figure}

@@

\begin{figure}
  \includegraphics[scale=0.7]{figureDifferentSearchModes.eps}
  \caption{Search Modes}
  \label{fgr:differentSearchModes}
\end{figure}

@@

Using a calibration or instrument noise percursor tolerance, such as 5ppm or 10ppm is a valid approach, since it identifies only exact matches.
A major drawback of this approach is that it misses monoisotope errors and co-isolated peptides.
To avoid those complications, and to get a more comprehensive list of matches, a wider search tolerance is used, e.g. $\pm 3.5$ Da.
This gives a substantially larger list of matches, and all the mass shifts could be attributed to co-isolation or missed monoisotopes.
A post-processing step may be employed to confidently identify the co-isolated and missed monoisotope matches, but a wide range of high-confidence matches with un-identified mass shifts remain.
These often correspond to modifications on the identified peptides, and require an extensive framework (such as G-PTM-D) to identify and label them.
More importantly, those modified peptides \textit{do not} correspond to the identified one, thus arguably defeating the purpose of a narrow-mass search.

We observe that when using an incomplete database that does not include comprehensive modification information, a notch search with notches at monoisotopic peaks is superior to either of the narrow-mass searches.

\begin{table}[]
\centering
\caption{Notch Search vs Narrow-Mass Search comparison of PSMs for Mouse}
\label{my-label}
\begin{tabular}{@{}llll@{}}
                    & 5ppm Search & 3.5 Da Search & Notch Search \\ \cmidrule(l){2-4}
PSMs Within 1\% FDR & 157188      & 178003        & 173483       \\
MisAssigned PSMs    & 0           & 7783          & 0            \\
Good PSMs           & 157188      & 170220        & 173483       \\ 
\end{tabular}
\end{table}

@@

\begin{figure}
\caption{Upper and Lower bound cutoffs in Mouse Data}
\label{fig:figure2-upperlowerbounds}
\includegraphics[scale=0.5]{figure_2-upperlowerbounds}
\end{figure}

@@

\begin{figure}
\caption{Some Histogram Peaks For Mouse Data}
\label{fig:figure3}
\includegraphics[scale=0.6]{figure_3peaks}
\end{figure}

>>>>

Correct identification of protein post-translational modifications (PTMs) is crucial to understanding many aspects of protein function in biological processes.
G-PTM-D\cite{Li_2016} is a recently developed tool for global identification and localization of PTMs.
Spectral file calibration prior to applying G-PTM-D, and algorithmic enhancements in the peptide database search shown here significantly increase the accuracy and efficiency of PTM identification.
We enhance G-PTM-D by using targeted mass-window searches, and demonstrate its effectiveness in identification of numerous types of PTMs, including high-mass modifications such as glycosylations.
The proposed changes lead to a 20\% increase in the number of identified modifications and an order of magnitude decrease in the search time.
Targeted mass-window searches are also shown to be useful as replacements to standard wide-mass searches, resulting in a 10\% increase in the overall number of confidently identified peptides, and a 5\% increase in the number of mass shifts observed.
Using a modification discovery heuristic originally based on the wide-mass search, targeted mass-window searches lead to an increase in the number of novel modifications discovered.
\end{abstract}

@@

\begin{figure}
  \includegraphics{diagram.eps}
  \caption{Workflows}
  \label{fgr:diagram}
\end{figure}

@@

\begin{figure}
  \includegraphics{fig2-calibrationQuality.eps}
  \caption{Histograms demonstrating the effect of calibration on PSM mass errors}
  \label{fgr:figure1}
\end{figure}

@@

\begin{figure}
\includegraphics{fig1-searchTypes.eps}
\caption{Search Modes}
\label{fig:fig1-searchtypes}
\end{figure}

@@

Using a calibration or instrument noise percursor tolerance, such as 5ppm or 10ppm is a valid approach, since it identifies only exact matches.
A major drawback of this strategy is that it misses monoisotope errors.
To avoid this complication, and to get a more comprehensive list of matches, a wider search tolerance is often used, e.g. $\pm 3.5$ Da.
This gives a substantially larger list of matches, and the resulting PSMs with non-zero mass shifts could be attributed to missed monoisotopes or co-isolated peptides.
While this approach provides a larger list of confident PSMs, a substantial fraction of those is wrong, since certain mass shifts in the chosen Dalton range arise due to sequence variations or PTMs.
These require an extensive framework (such as G-PTM-D) to identify and label them.
More importantly, those modified peptides \textit{do not} correspond to the identified one, thus arguably defeating the purpose of a narrow-mass search.

We observe that when using an incomplete database that does not include comprehensive modification and sequence variation information, a notch search with notches at monoisotopic peaks is superior to either of the narrow-mass searches.

\begin{table}[]
\centering
\caption{Notch Search vs Narrow-Mass Search comparison of PSMs for Mouse}
\label{my-label}
\begin{tabular}{@{}llll@{}}
                    & 5ppm Search & 3.5 Da Search & Notch Search \\ \cmidrule(l){2-4}
PSMs Within 1\% FDR & 157188      & 178003        &        \\
MisAssigned PSMs    & 0           &               & 0            \\
Good PSMs           & 157188      &         &        \\ 
\end{tabular}
\end{table}

\begin{figure}
\caption{Mouse histograms around 0}
\label{fig:fig3mouse-1012}
\includegraphics{fig3mouse-1012}
\end{figure}

\begin{figure}
\caption{Jurkat histograms around 0}
\label{fig:fig3jurkat-1012}
\includegraphics{fig3jurkat-1012}
\end{figure}

@@

\begin{figure}
\caption{Upper and Lower bound cutoffs in Mouse Data}
\label{fig:fig4-limitsOnOpenSearch}
\includegraphics{fig4-limitsOnOpenSearch}
\end{figure}

@@

\begin{figure}
\caption{Some Histogram Peaks For Mouse Data}
\label{fig:fig5-HistogramsAround80and113}
\includegraphics{fig5-HistogramsAround80and113}
\end{figure}
